\texttt{MSnbase}, efficient and elegant R-based processing and visualisation of raw mass spectrometry data

### 2020-07-23 msnbase2.Rnw
rename section Package maintenance and governance
define elegant
explain subsetting overhead - closes #16

In the following sections, we describe \sout{\texttt{MSnbase}'s
  ability to efficiently handle very large mass spectrometry data and
  experiments and how it} \textcolor{red}{how \texttt{MSnbase}} can be
used for data processing and visualisation. \textcolor{red}{An example
  of its ability to efficiently handle also very large mass
  spectrometry experiments (in this case with 5,773,464 spectra in
  1,182 mzXML files) is provided as supplementary information.}  We
will also illustrate how it makes use of the forward-pipe operator
(\texttt{\%>\%}) defined in the \texttt{magrittr} package. This
operator has proved useful to develop non-trivial analyses by
combining individual functions into easily readable pipelines.

@@

\begin{figure}[p]
  \centering
<<plot_bench>>=
(p_read + p_sz + p_filt) / p_access +
    plot_layout(heights = c(1, 0.75)) +
    plot_annotation(tag_levels = 'a') &
    theme(axis.title = element_text(size = 10))
@
\caption{(a) Reading time (in seconds) and (b) data size in memory (in
  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
  only) and 6103 MS2 (on-disk and in-memory) spectra. (c) Filtering
  benchmark assessed over 10 interactions on in-memory and on-disk
  data containing 6103 MS2 spectra.  (d) Access time to spectra for
  the in-memory (left) and on-disk (right) backends for 1, 10, 100
  1000, 5000 and all 6103 spectra. Benchmarks were performed on a Dell
  XPS laptop with an Intel i5-8250U processor 1.60 GHz (4 cores, 8
  threads), 7.5 GB RAM running Ubuntu 18.04.4 LTS 64-bit
  \textcolor{red}{and SSD drive}. }
\label{fig:bench}
\end{figure}

@@

As shown on Figure~\ref{fig:bench} (c), this lazy mechanism is
significantly faster than its application on in-memory data. The
advantageous reading and execution times and memory footprint of the
on-disk backend are possible by retrieving only spectra data from the selected
subset hence avoiding access to the full raw
data. Once access to the spectra m/z and intensity values become
mandatory (for example for plotting), then the in-memory backend
becomes more efficient, as illustrated on Figure~\ref{fig:bench}
(d). This gain is maximal when the whole dataset is accessed
(i.e. all spectra are already in memory) and negligible when only subsets of the
data are requested.

@@

\subsection{\textcolor{red}{Package management and governance}}

>>>>

In the following sections, we describe \sout{\texttt{MSnbase}'s
  ability to efficiently handle very large mass spectrometry data and
  experiments and how it} \textcolor{red}{how \texttt{MSnbase}} can be
used for data processing and visualisation. \textcolor{red}{An example
  of its ability to efficiently handle also very large mass
  spectrometry experiments (in this case with 5,773,464 spectra in
  1,182 mzXML files) is provided as supplementary information.}  We
will also illustrate how it makes use of the forward-pipe operator
(\texttt{\%>\%}) defined in the \texttt{magrittr} package. This
operator has proved useful to develop non-trivial analyses by
combining individual functions into easily readable
\textcolor{red}{and elegant} pipelines.

@@

\begin{figure}[p]
  \centering
<<plot_bench>>=
(p_read + p_sz + p_filt) / p_access +
    plot_layout(heights = c(1, 0.75)) +
    plot_annotation(tag_levels = 'a') &
    theme(axis.title = element_text(size = 10))
@
\caption{(a) Reading time (in seconds) and (b) data size in memory (in
  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
  only) and 6103 MS2 (on-disk and in-memory) spectra. (c) Filtering
  benchmark assessed over 10 interactions on in-memory and on-disk
  data containing 6103 MS2 spectra.  (d) Access time to spectra for
  the in-memory (left) and on-disk (right) backends for 1, 10, 100
  1000, 5000 and all 6103 spectra. Benchmarks were performed on a Dell
  XPS laptop with an Intel i5-8250U processor 1.60 GHz (4 cores, 8
  threads), 7.5 GB RAM running Ubuntu 18.04.4 LTS 64-bit
  \textcolor{red}{and an SSD drive}. }
\label{fig:bench}
\end{figure}

@@

As shown on Figure~\ref{fig:bench} (c), this lazy mechanism is
significantly faster than its application on in-memory data. The
advantageous reading and execution times and memory footprint of the
on-disk backend are possible by retrieving only spectra data from the
selected subset hence avoiding access to the full raw data. Once
access to the spectra m/z and intensity values become mandatory (for
example for plotting), then the in-memory backend becomes more
efficient, as illustrated on Figure~\ref{fig:bench} (d).
\textcolor{red}{The benefit of accessing data in memory is however
  reduced by underlying copies that are performed during the
  subsetting operation. When subsetting an in-memory \texttt{MSnExp}
  into a new, smaller in-memory \texttt{MSnExp} instance, the matrices
  that contain the spectra for the new object are copied, thus leading
  to increased execution time and (transient, if the original data is
  replaced) memory usage. Figure~\ref{fig:bench} (d) shows that the
  larger the subset, that smaller the benefits of an in-memory
  backend. The example with the 6103 spectra, corresponding to the
  full data (i.e. all spectra are already in memory and there is no
  memory management overhead) is representative of memory access only
  and constitutes the best case scenario. } The gain of an in-memory
backend is maximal when the whole dataset is accessed \sout{(i.e. all
  spectra are already in memory)} and becomes negligible when large
subsets of the data are requested.

@@

\subsection{\textcolor{red}{Package maintenance and governance}}
